---
title: "Machine Learning on the RECS 2009 Dataset"
output: html_notebook
---

#### Background & Business Value
Why predict residential kWh usage?
- Description (features that correlate most strongly with use)
- Prediction (new homeowner for cost forecasting; policy forecasting usage in residential sector)

#### EDA and Feature Engineering
1) Pre-process
- Pre-processing of variable codebook (ensuring correct variable types numeric/integer/factor)
- Remove major missingness (< 5% complete)
- Remove obvious collinearity among a few variables
- Remove dependent variables other than KWH (e.g. elec usage in BTU)
- Remove imputation flag variables added by EIA

2) Dimension reduction
- Still hundreds of variables after pre-processing
- Pick out numeric variables and run correlation plot on them (still some obvious collinearity)
```{r}
# cor.plot
```
- PCA reveals most numeric variability picked up within 10-15 components
```{r}
# PCA plot
```
- LASSO regression to see regularization to of many less-predictive variables; drop most vars which go to 0

#### Modeling 
3) Regression - prediction of KWH
- GLM with grid search
- GBM and RF with grid search
- Variable importances in common
```{r}
# Variable importance plots faceted by model type
# Scatter plots with predicted v. actual for most important vars
# Table with train/test RMSE values comparing performace
```

