---
title: "How @realDonaldTrump tweeted"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Background and data

@realDonaldTrump probably won't be used as heavily as it has been since 2009, when Doanld J. Trump registered his Twitter account and started building his online presence. (Now in case you are interested, you should be following @POTUS.) However, the entire database of his tweets going back to 2009 has been archived and made searchable at <http://trumptwitterarchive.com/>. This project serves as a preliminary exploration of some of Trump's tweeting patterns, with a focus on identifying the elements that attracted Twitter users' attention throughout his 2016 presidential campaign. 

The data was downloaded from <https://github.com/bpb27/political_twitter_archive/blob/master/realdonaldtrump/realdonaldtrump.csv>:

```{r, echo=FALSE}
data <- read.csv('data/realdonaldtrump_0122.csv',header=T, stringsAsFactors = F)
```
```{r, echo=TRUE}
summary(data)
```

## Including Plots

You can also embed plots, for example:

```{r ggplot2 dplyr, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
